---
title: "Logistic Regression"
date: today
# For author options see : 
# https://quarto.org/docs/authoring/front-matter.html#authors-and-affiliations
# NB: multiple authors can be added here.
author:
  - name:
      given: Joshua
      family: Wilson Black
    email: joshua.black@canterbury.ac.nz
    orcid: 0000-0002-8272-5763
    affiliation: 
      - "Te Kāhui Roro Reo | New Zealand Institute of Language, Brain and Behaviour"
      - "Te Whare Wānanga o Waitaha | University of Canterbury"
format:
  revealjs:
    theme: [custom.scss]
    incremental: true
    logo: images/NZILBB-small.svg
    df-print: paged
    template-partials:
      - title-slide.html
    title-slide-attributes:
      # If you have Marsden funding, change image name to `nzilbb-uc-marsden.svg`
      data-background-image: images/nzilbb-uc.svg
      # First number controls the horizontal position, second controls vertical.
      data-background-position: '50% 5%'
      # Controls size of image relative to width of the slide.
      data-background-size: 50%
    embed-resources: false
    include-in-header:
      - text: |
          <link rel="icon" type="image/png" sizes="32x32" href=".//images/fav.png" />
bibliography: 
  - grateful-refs.bib
  - stat_workshops.bib
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: true
knitr:
  opts_chunk: 
    dev: "png"
    dev.args: 
      bg: "transparent"
---

```{r}
#| echo: false
library(tidyverse)
library(glue)
library(here)
library(lme4)
library(lmerTest)
library(modelbased)
library(patchwork)
library(ModelMetrics)

theme_set(theme_minimal(base_size = 20))
```

## Overview

1. What is a logistic regression?
    - ...and generalised linear models more broadly
2. Examples of logistic and mixed-effects logistic regressions in R.

## About you

- You know something about linear regression.
- But, how can I use linear models for non-linear explanatory variables!?
    - e.g. presence or absence of a linguistic variant.
    - e.g. binary choice in an experiment.

## Crash course

::: r-stack

:::: fragment

```{r}
#| echo: false
plot_data <- 
  tibble(
    x = rnorm(n=1000),
    y = x * 2.5 + rnorm(n=1000, sd = 3)
  )

plot_data |> 
  ggplot(
    aes(
      x = x,
      y = y
    )
  ) +
  geom_point(alpha = 0.7)
```


::::

:::: fragment

```{r}
#| echo: false
plot_data |> 
  ggplot(
    aes(
      x = x,
      y = y
    )
  ) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", linewidth = 2, colour = "blue", se = FALSE)
```



::::

:::: fragment

[Draw straight line through dots!]{.big .backlight}

::::

:::: fragment

```{r}
#| echo: false
plot_data <- 
  tibble(
    x = rnorm(n=1000),
    y = if_else(
        x < 0, 
        rbinom(n=1000, size=1, prob=0.3),
        rbinom(n=1000, size=1, prob=0.8)
    )
  ) |> 
  mutate(
    y = factor(
      if_else(y == 0, "Variant 1", "Variant 2"),
    )
  )

plot_data |> 
  ggplot(
    aes(
      x = x,
      y = y
    )
  ) +
  geom_jitter(height = 0.01, alpha = 0.3) +
```

::::

:::: fragment

[🤷]{.real-big}

::::

:::: fragment

```{r}
#| echo: false
plot_data |>
  mutate(
    y = if_else(
      y == "Variant 1", 0, 1
    )
  ) |>
  ggplot(
    aes(
      x = x,
      y = y
    )
  ) +
  geom_jitter(height = 0.01, alpha = 0.3)
```

::::

:::: fragment

```{r}
#| echo: false
plot_data |>
  mutate(
    y = if_else(
      y == "Variant 1", 0, 1
    )
  ) |>
  ggplot(
    aes(
      x = x,
      y = y
    )
  ) +
  geom_jitter(height = 0.01, alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 2)
```

::::

:::: fragment

[🙅]{.real-big}

::::

:::: fragment

```{r}
#| echo: false
plot_data |>
  mutate(
    y = if_else(
      y == "Variant 1", 0, 1
    )
  ) |>
  ggplot(
    aes(
      x = x,
      y = y
    )
  ) +
  geom_jitter(height = 0.01, alpha = 0.3) +
  geom_smooth(method = "glm", method.args = "binomial", se = FALSE, linewidth = 2)
```

::::

:::: fragment

[✨🧙✨]{.real-big}

::::

:::

## Generalised linear models

- We need: something associated with our explanatory variable which we 
can model using the tools of linear regression.
- [Generalised linear models]{.red} are a group of models which enable this.
- [Logistic regression]{.red} is a generalised linear model which models the
'log-odds' (= 'logits') of an event.

## Odds {.smaller}

- [🤖]{.big}: _The odds of successfully navigating an asteroid field are 3720 to 1_
    - i.e.: in 3721 trials, you'll succeed once.
    - i.e.: the probability is 1/3721 = 0.0003.
- [🎭]{.big}: _Knew that we ventured on such dangerous seas_  
_That if we wrought out life 'twas ten to one_
    - i.e.: if we go to see 11 times, we'll die 10 times.
    - i.e.: probability of survival is 1/11 = 0.0909...
- Odds from probability: $\frac{p}{1-p}$
    
## Log Odds

:::: fragment
```{r}
#| echo: false
plot_data <- 
  tibble(
    x = seq(0, 0.999, by = 0.01)
  ) |> 
  mutate(
    y = x
  )

annote_points <- tibble(
  prob = c(0.1, 0.5, 0.9),
  odds = prob/(1-prob),
  log_odds = log(odds),
  label_prob = as.character(glue("Prob={prob}")),
  label_odds = as.character(
    glue("{prob}/(1-{prob})={round(odds, 2)}")
  ),
  label_logodds = as.character(
    glue(
      "log({round(odds, 2)})={round(log_odds, 2)} "
    )
  )
)

prob_plot <- plot_data |> 
  ggplot(
    aes(
      x = x,
      y = y
    )  
  ) +
  geom_path(linewidth = 1) +
  geom_point(aes(x=prob, y=prob), size = 2, data=annote_points) +
  labs(
    y = "Probability",
    x = "Probability"
  )

odds_plot <- plot_data |>
  ggplot(
    aes(
      x = x,
      y = x/(1-x)
    )
  ) +
  geom_path(linewidth = 1) +
  geom_point(aes(x=prob, y=odds), size = 2, data=annote_points) +
  labs(
    y = "Odds (p/(1-p))",
    x = "Probability"
  )

log_plot <- plot_data |> 
  ggplot(
    aes(
      x = x,
      y = log(x/(1-x))
    )
  ) +
  geom_path(linewidth = 1) +
  geom_point(aes(x=prob, y=log_odds), size = 2, data=annote_points) +
  labs(
    y = "Log odds",
    x = "Probability"
  )

prob_plot + odds_plot + log_plot
```
::::

- Prob: 0 to 1; Odds: 0 to +∞; Log odds: -∞ to +∞ 

## Regression with Log-Odds

```{r}
#| echo: false
set.seed(2)
sim_data <- 
  tibble(
    cool_points = rnorm(n=70)
  ) |> 
  mutate(
    prob = (cool_points-min(cool_points))/
      (max(cool_points)-min(cool_points)),
    y_binom = map_dbl(prob, ~ rbinom(n=1, size=1, prob=.x)),
    variant = factor(
      if_else(y_binom == 0, "Variant 1", "Variant 2")
    )
  )

sim_data |> 
  ggplot(
    aes(
      x = cool_points,
      y = variant
    )
  ) +
  geom_jitter(height = 0.1)
```

## Regression with Log-Odds (2)

- Use the `glm()` function with `family = binomial`.

::: fragment
```{r}
fit_1 <- glm(variant ~ cool_points, data = sim_data, family = binomial) 
```
:::

## Model Diagnostics

- Standard residual plots aren't very informative!

::: fragment
```{r}
#| echo: false
to_plot <- 
  tibble(
    linear_pred = predict(fit_1),
    residuals = resid(fit_1)
  )

to_plot |> 
  ggplot(
    aes(
      x = linear_pred,
      y = residuals
    )
  ) +
  geom_point()
```

::::

## Model Diagnostics (2) {.smaller}

- One possibility is to 'bin' the residuals like so:

::: fragment

```{r}
#| echo: false
to_plot <- to_plot |> 
  mutate(
    linear_bins = cut(linear_pred, breaks = 20, labels = FALSE)
  ) |> 
  group_by(linear_bins) |> 
  summarise(
    residuals = mean(residuals),
    n = n()
  )

to_plot |> 
  ggplot(
    aes(
      x = linear_bins,
      y = residuals,
      size = sqrt(n)
    )
  ) +
  geom_point()
```

:::

- For more see advice on readings at end of slides.

## Goodness of Fit

- AUC is a good method, which is not uncommon.
- AUC measures classification performance.

::: fragment
```{r}
#| output-location: fragment
# auc() is from ModelMetrics.
auc(fit_1)
```
:::

- AUC of 0.5 is random guessing.
- Report some measure of goodness of fit.

## Model Coefficients {.smaller}

```{r}

broom::tidy(fit_1)
```

- Cool points is scaled so `(Intercept)` gives log-odds of mean `cool_points`. 
  - i.e.: more likely to use 'Variant 2'.
  - How much? Prob = `plogis(0.35)` = 0.59.
- Each 1 s.d. increase in `cool_points` increases log odds by 1.
  - This means different things at different levels of _probability_ 

## Plot Predictions (response)

::: fragment

```{r}
#| warning: false
#| output-location: fragment
# Using estimate_relation from modelbased package.
estimate_relation(
    fit_1,
    by = "cool_points"
  ) |> 
  plot()
```

:::

## Plot Predictions (link)

::: fragment

```{r}
#| warning: false
#| output-location: fragment
estimate_link(
    fit_1,
    by = "cool_points"
  ) |> 
  plot()
```

:::

## When?

- Any time you have a _binary_ distinction.
- Forced choice in experiments.
- A linguistic variant of interest vs. all other variants.
- More than two categories? Maybe try [multinomial logistic regression](https://bookdown.org/chua/ber642_advanced_regression/multinomial-logistic-regression.html).

## Bells and Whistles

- Fit mixed effects with `glmer()` and `family = binomial`.

```{r}
#| echo: false
# Create a new example.
sim_participants <-  
  tibble(
    participant_name = randomNames::randomNames(70),
    year_of_birth = sample(seq(1920:1999), size = 70, replace = T),
    rides_bike = sample(c(TRUE, FALSE), size = 70, replace = T),
    cool_points = if_else(rides_bike, rnorm(n=70), rnorm(n=70, mean=0.1)),
    eats_pasta = sample(c(TRUE, FALSE), size = 70, replace = T),
    idiosyncracy = rnorm(n=70, sd=0.5)
  ) 

sim_obs <- sim_participants |> 
  mutate(
    more_explanatory = map(
      participant_name,
      ~ {
        n_obs = sample(10:60, size = 1)
        tibble(
          preceeding_environment = factor(
            sample(
              c("env_1", "env_2", "env_3"), size = n_obs, replace = T
            )
          ),
          following_environment = factor(
            sample(
              c("env_1", "env_2", "env_3"), size = n_obs, replace = T
            )
          ),
          style = factor(
            sample(
              c("formal", "informal"), size = n_obs, replace = T
            )
          ),
        )
      }
    )
  ) |> 
  unnest(more_explanatory) |> 
  mutate(
    pe_coef = case_when(
      preceeding_environment == "env_1" ~ -0.2,
      preceeding_environment == "env_2" ~ 0,
      preceeding_environment == "env_3" ~ 0.1,
    ),
    fe_coef = case_when(
      preceeding_environment == "env_1" ~ 0.1,
      preceeding_environment == "env_2" ~ -0,
      preceeding_environment == "env_3" ~ -0.25,
    ), 
    style_coef = if_else(
      style == "formal", -0.9, 0.2
    ),
    yob_s = scale(year_of_birth),
    log_odds = 
      -0.2 + 0.01 * year_of_birth + 0.5*rides_bike + 0.3*cool_points + pe_coef + 
      fe_coef + style_coef + idiosyncracy,
    prob = plogis(log_odds),
    y_binom = map_dbl(prob, ~ rbinom(n=1, size=1, prob=.x)),
    variant = factor(if_else(
      y_binom == 0, "Variant 1", "Variant_2"
    ))
  )

```


::: fragment

```{r}
fit_2 <- glmer(
  variant ~ year_of_birth + cool_points + rides_bike + eats_pasta + style +
    (1|participant_name),
  data = sim_obs,
  family = binomial
)
```

:::

## Summary

```{r}
summary(fit_2)
```
## Plot Predictions

```{r}
estimate_relation(
    fit_2,
    by= c("cool_points", "rides_bike")
  ) |> 
  plot()
```


## Further Reading

- @sondereggerRegressionModelingLinguistic2023, Chapter 6.
- @winterStatisticsLinguistsIntroduction2019, Chapter 12.

# References

```{r}
#| echo: false
grateful::nocite_references(
  grateful::cite_packages(output = "citekeys", out.dir = here())
)
```


::: refs

:::
